Follow-up Samples

### 2017-04-21 submission/manuscript.tex
re-render good version

\section{The Fecal Microbiome Before and After Treatment for Colorectal
Adenoma or
Carcinoma}\label{the-fecal-microbiome-before-and-after-treatment-for-colorectal-adenoma-or-carcinoma}

\vspace{25mm}

\begin{center}
Running Title: Human Microbiome before and after Colorectal Cancer

\vspace{10mm}

@@

\textbf{Background.} Colorectal cancer (CRC) is a worldwide health
problem. Despite growing evidence that members of the gut microbiota can
drive tumorigenesis, little is known about what happens to the
microbiome after treatment for an adenoma or carcinoma. This study
tested the hypothesis that treatment for adenoma or carcinoma alters the
abundance of bacterial populations associated with with disease to those
associated with a normal colon. We tested this hypothesis by sequencing
the 16S rRNA genes in the feces of 67 individuals before and after
treatment for adenoma (N=22), advanced adenoma (N=19), and carcinoma
(N=26).

\textbf{Results.} There were large changes the the bacterial communities
associated with treatment across the three groups. The communities from
patients with carcinomas changed significantly more than those with
adenoma following treatment (P-value=5.4e-05); however, there was not a
significant difference between those with advanced adenoma and those
with adenoma or carcinoma (P-value\textgreater{}0.05). Although
treatment brought about large intrapersonal changes, the change in the
abundance of individual OTUs to treatment was not consistent within
diagnosis groups (P-value \textgreater{} 0.05). Because the distribution
of OTUs across patients and diagnosis groups was patchy, we used the
Random Forest machine learning algorithm to identify groups of OTUs that
allowed us to successfully distinguish between pre- and post-treatment
samples for each of the diagnosis groups. However, across the three
models, there was little overlap between the OTUs that were indicative
of treatment. Next, we used a larger cohort that contained individuals
with normal colons and those with adenomas, advanced adenomas, and
carcinomas to determine whether individuals who underwent a treatment
were more likely to have OTUs associated with normal colons. We again
built Random Forest models and measured the change in the positive
probability of having one of the three diagnoses. Although we could
clearly differentiate pre- and post-treatment communities from the three
diagnosis groups, only those patients that initially had carcinomas
experienced a significant decrease in positive probability of having a
carcinoma (P-value \textless{} 0.05). Finally, tested whether the type
of treatment impacted the microbiota of those diagnosed with carcinomas
and were unable to detect any significant differences in characteristics
of these communities between individuals treated with chemotherapy and
those treated with chemotherapy and radiation (P-value \textgreater{}
0.05).

\textbf{Conclusions.} Further exploration of the relationship between
diagnosis, treatment, and the impact on the microbiome will yield
improvements in disease management.

@@

Building on this evidence, several human-based studies have identified
unique signatures of colonic lesions {[}14--19{]}. One line of research
has identified community-level differences between those bacteria that
are found on and adjacent to colonic lesions and have supported a role
for \emph{Bacteroides fragilis}, \emph{Escherichia coli}, and
\emph{Fusobacterium nucleatum} in tumorigenesis {[}20--22{]}. Others
have proposed feces-based biomarkers that could be used to diagnose the
presence of colonic adenomas and carcinomas {[}23--25{]}. These studies
have associated \emph{Fusobacterium nucleatum} and other oral pathogens
with lesions. They have also noted that the loss of bacteria generally
thought to produce short chain fatty acids, which can suppress
inflammation, is associated with colonic lesions. Gut bacteria have a
role in tumorigenesis and there is promise that these populations may be
useful biomarkers for aiding in the early detection of disease.

@@

\textbf{\emph{Treatment alters the bacterial community structure of
patients diagnosed with colonic lesions.}} Within our 67-person cohort
we tested whether the microbiome of patients with adenoma (N=22),
advanced adenoma (N=19), or carcinoma (N=26) had any broad differences
between pre- and post-treatment samples (Table 1). The structure of the
microbial communities of the pre- and post-treatment samples differed
{[}Figure 1A{]}. We found that the communities obtained pre- and
post-treament among the patients with carcinomas changed significantly
more than those patients with adenoma (P-value \textless{} 0.001). There
were no significant differences in the amount of change observed between
the patients with adenoma and advanced adenoma or between the patients
with advanced adenoma and carcinoma (P\textgreater{}0.05). Next, we
tested whether there was a consistent direction in the change in the
community structure between the pre and post-treatment samples for each
of the diagnosis groups {[}Figure 1BCD{]}. We only observed a consistent
shift in community structure for the patients with carcinoma when using
a PERMANOVA test (adenoma P-value0.999, advanced adenoma P-value0.945,
and carcinoma P-value0.005). Finally, we measured the number of observed
OTUs, Shannon evenness, and Shannon diversity in the pre and
post-treatment samples and did not observe a significant change for any
of the diagnosis groups (P-value\textgreater{}0.05) {[}Table S1{]}.

\textbf{\emph{The effects of treatment are not consistent across
treatment groups.}} We used two approaches to identify those bacterial
populations that change between the two samples for each diagnosis
group. First, we sought to identify individual OTUs that could account
for the change in overall community structure. However, using a paired
Wilcoxon test we were unable to identify any OTUs that were
significantly different in the pre and post-treatment groups {[}Figure
S1{]}. It is likely that high inter-individual variation and the patchy
distribution of OTUs across individuals limited the statistical power of
the test. To overcome these problems we developed Random Forest models
to identify collections of OTUs that would allow us to differentiate
between pre and post-treatment samples from each of the diagnosis
groups. Because of the relatively small number of subject in each group,
we restricted our models to only incorporate 10 OTUs to limit the
likelihood that that models would overfit the data. Despite this
restriction, the models performed well (adenoma AUC=0.69 - 0.92,
advanced adenoma AUC=0.8 - 1, carcinoma AUC=0.82 - 0.98). Interestingly,
the 10 OTUs that were used for each model had little overlap with each
other {[}Figure 2{]}. These results support the earlier community-wide
analysis where we observed that the treatment had an impact on the
overall community structure; however, the effect of treatment was not
consistent across patients and diagnosis groups.

\textbf{\emph{Post-treatment samples from patients with carcinoma
share}} Next, we determined whether treatment changed the microbiota in
a way that the post-treatment communities resembled that of patients
with normal colons. To test this, we used an expanded cohort of 423
individuals that were diagnosed under the same protocol as having normal
colons or colons with adenoma, advanced adenoma, or carcinoma (Table 2).
We then constructed Random Forest models to classify the pre and
post-treatment samples as having their original diagnosis or having a
normal colon. The models performed well (adenoma AUC=0.62 - 0.72,
advanced adenoma AUC=0.68 - 0.77, carcinoma AUC=0.84 - 0.9; Figure S2).
The OTUs that were incorporated into the adenoma and advanced adenoma
models largely overlapped and those OTUs that were used to classify the
carcinoma samples were largely distinct from those of the other two
models {[}Figure 4A{]}. Among the OTUs that were shared across the three
models were those populations commonly considered as commensals (e.g.
\emph{Faecalibacterium}, \emph{Lachnospiraceae}, \emph{Bacteroides},
\emph{Dorea}, \emph{Anaerostipes}, and \emph{Roseburia}) {[}Figures
4B{]}. Although many of these OTUs were also included in the model
differentiating between patients with normal colons and those with
carcinoma, this model also included OTUs affiliated with populations
that have previously been associated with carcinoma
(\emph{Fusobacterium}, \emph{Porphyromonas}, \emph{Parvimonas})
{[}Figure S3{]}. Finally, we applied these three models to the pre and
post-treatment samples for each diagnosis group and quantified the
change in the positive probability of the model. A decrease in the
positive probability would indicate that the colon more closely
resembled that of a patient with a normal colon. There was not a
significant change in the positive probability for the adenoma or
advanced adenoma groups {[}Figure 3{]}. The positive probability for the
pre and post-treatment samples from patients diagnosed with carcinoma
significantly decreased. In fact, only 6 of the 26 patients (23.08\%)
that were initially diagnosed with a carcinoma had a higher positive
probability after treatment. One of those was re-diagnosed with
carcinoma on the follow up visit. These results indicate that although
there are changes in the microbiota associated with treatment, those
experienced by the patients in the carcinoma group were the only ones
where the change was directed towards what would be found in a normal
colon

@@

This study builds upon previous work that demonstrated a role for the
microbiota in tumorigenesis and that the microbiome contains biomarkers
to diagnosis colonic lesions {[}5--9,23--27{]}. Our study focused on
comparing the microbiome of patients diagnosed with adenoma, advanced
adenoma, and carcinoma before and after treatment. For all three groups
of patients we observed changes in their microbiota. The microbiota of
patients diagnosed with carcinoma changed significantly more than the
other groups. This change resulted in communities that more closely
resembled those of patients with normal colons.

For the carcinoma group there was a larger change between their initial
and follow up samples then for either adenoma or advanced adenoma. This
change was directionaly the same for each individual with carcinoma.
This observed change has important implications for future research into
CRC recurrence. Knowing that the community changes towards normal after
treatment means that it is reasonable to hypothesize that the microbiome
could be a good biomarker for stratifying individuals between those most
and least at risk for CRC recurrence. Interestingly, our 23.08\%
prevalence of increased carcinoma probability fits nicely within the
range quoted for CRC recurrence (20-30\%) {[}3,4{]}.

A potential limitation, within our study, is that we do not know whether
individuals who had an increased positive probability for carcinoma on
follow up eventually had a subsequent CRC recurrence. This information
would help to strengthen the case for the ability of our model to be
used as a disease stratification tool for CRC recurrence. This would be
an exciting area of future research that would be able to link treatment
changes within the microbiome and CRC recurrence together.

The microbiome changes associated with carcinoma were found to be
towards those with normal colons. This would suggest that treatment for
CRC is not only successful for removing the carcinoma but also at
removing bacteria associated with them. This is important due to the
numerous studies that have pointed towards a link between the community
and tumorgenesis {[}10--12{]}. The importance of the community to this
process is further highlighted in our data {[}Figure 2, 4, \& S3{]}.
Interestingly, those who had an increase in their positive probability
of carcinoma had no change in their median value for oral pathogens
found within the carcinoma model but had a decrease in non-oral
pathogens {[}Figure S4{]}. Although this finding was not statistically
significant, this trend would further highlight the importance of
community to tumorgenesis. Future studies with larger power may be able
to confirm these reported observations.

Although we were not able to identify specific OTUs due to the
heterogeneity of the population. This is a common problem that occurs in
other fields of microbiome study {[}28{]}. This made it very difficult
to identify individual OTUs associated with treatment and recovery
towards a normal colon. To overcome this large individual heterogeneity
we used RF models that made it easier to identify consortia of
microbiota that were associated with treatment and recovery. A potential
future direction would be to study whether specific gene signatures
associated with these consortia of bacteria have a role to play in the
recurrence of CRC.

Despite some of these stated shortcomings our findings add to the
existing scientific knowledge on CRC and the microbiome: That there is a
measurable change in the bacterial community after adenoma, advanced
adenoma, or carcinoma treatment. Further, individuals with carcinoma had
a larger change than adenoma or advanced adenoma indivdiuals and this
change in community was towards those with normal colons. Our data
provides additional evidence on the importance of the community on
recovery of the microbiome after CRC treatment and opens interesting
avenues of research into how these changes may affect recurrence.

@@

\textbf{\emph{Study Design and Patient Sampling:}} Sampling and design
have been previously reported in Baxter, et al {[}23{]}. Briefly, study
exclusion involved those who had already undergone surgery, radiation,
or chemotherapy, had colorectal cancer before a baseline fecal sample
could be obtained, had IBD, a known hereditary non-polyposis colorectal
cancer, or familial adenomatous polyposis. Samples used to build the
models for prediction were collected either prior to a colonoscopy or
between 1 - 2 weeks after. The bacterial community has been shown to
normalize back to a pre-colonoscopy community within this time period
{[}29{]}. Our training cohort consisted of a total of 423 individuals
{[}Table 1{]}. Our study cohort consisted of 67 individuals with an
initial sample as described and a follow up sample obtained between 188
- 546 days after treatment of lesion {[}Table 2{]}. This study was
approved by the University of Michigan Institutional Review Board. All
study participants provided informed consent and the study itself
conformed to the guidelines set out by the Helsinki Declaration.

\textbf{\emph{16S rRNA Gene Sequencing:}} Sequencing was completed as
described by Kozich, et al. {[}30{]}. DNA extraction used the 96-well
Soil DNA isolation kit (MO BIO Laboratories) and an epMotion 5075
automated pipetting system (Eppendorf). The V4 variable region was
amplified and the resulting product was split between three sequencing
runs with normal, adenoma, and carcinoma evenly represented on each run.
Each group was randomly assigned to avoid biases based on sample
collection location. The pre- and post-treatment samples were sequenced
on the same run.

\textbf{\emph{Sequence Processing:}} The mothur software package
(v1.37.5) was used to process the 16S rRNA gene sequences and has been
previously described {[}30{]}. The general workflow using mothur was:
Paired-end reads were first merged into contigs, quality filtered,
aligned to the SILVA database, screened for chimeras, classified with a
naive Bayesian classifier using the Ribosomal Database Project (RDP),
and clustered into Operational Taxonomic Units (OTUs) using a 97\%
similarity cutoff with an average neighbor clustering algorithm. The
number of sequences for each sample was rarefied to 10523 to minimize
uneven sampling.

\textbf{\emph{Model Building:}} The Random Forest {[}31{]} algorithm was
used to create the model used to create the three models used. The
adenoma model classified normal versus adenoma, advanced adenoma was
normal versus advanced adenoma, and carcinoma was normal versus
carcinoma. The toal number of individuals in this data set was 423
individuals. There were a total of 239 individuals in the adenoma model,
262 individuals in the advanced adenoma model, and 266 indivdiuals in
the carcinoma model {[}Table 1{]}. Each model was then applied to our
67-person cohort testing prediction of adenoma intial (adenoma n = 22)
versus adenoma follow up (adenoma n = 0), advanced adenoma initial
(advanced adenoma n = 19) versus advanced adenoma follow up (advanced
adenoma n = 0), carcinoma initial (carinoma n = 26) versus carcinoma
follow up (carcinoma n = 1).

@@

Assessment of the most important OTUs to the model involved counting the
number of times an OTU was present in the top 10\% of mean decrease in
accuracy (MDA) for each of the 100 different splits run. This was then
followed with filtering of this list to variables that were only present
in more than 50\% of these 100 runs. The final collated list of
variables was then run through the mtry optimization again. Once the
ideal mtry was found the entire sample set specific to normal versus
adenoma, normal versus advanced adenoma, or normal versus carcinoma was
used to create the final Random Forest model on which classifications on
the 67-person cohort was completed. For all three models the final
optimized mtry was 2

The default cutoff of 0.5 was used as the threshold to classify
individuals as positive or negative for lesion. The hyper-parameter,
mtry, defines the number of variables to investigate at each split
before a new division of the data was created with the Random Forest
model.

\textbf{\emph{Statistical Analysis:}} The R software package (v3.3.2)
was used for all statistical analysis. Comparisons between bacterial
community structure utilized PERMANOVA {[}32{]} in the vegan package
(v2.4.1). Comparisons between probabilities as well as overall OTU
differences between pre- and post-treatment samples utilized a paired
Wilcoxson ranked sum test. Where multiple comparison testing was
appropriate, a Benjamini-Hochberg (BH) correction was applied {[}33{]}
and a corrected P-value of less than 0.05 was considered significant.
Unless otherwise stated the P-values reported are those that were BH
corrected. Model rank importance was determined by obtining the median
MDA from the 100 20 repeated 10-fold cross validation and then ranking
from largest to smallest MDA.

\textbf{\emph{Analysis Overview:}} We first tested whether there were
any differences between pre- and post-treatment samples in alpha and
beta diversity based on adenoma, SRN, or carcinoma. We then tested all
OTUs for differences between pre- and post-treatment samples. We next
used our specific models for adenoma, SRN, and carcinoma to test
classification accuracy, response towards a normal microbiome, and
common OTUs used across models. Finally, for the adenoma group
differences between those that received surgery or not was tested while
for the carcinoma group differences between those receiving chemotherapy
and radiation was tested.

\textbf{\emph{Reproducible Methods:}} A detailed and reproducible
description of how the data were processed and analyzed can be found at
\url{https://github.com/SchlossLab/Sze_followUps_2017}. Raw sequences
have been deposited into the NCBI Sequence Read Archive (SRP062005 and
SRP096978) and the necessary metadata can be found at
\url{https://www.ncbi.nlm.nih.gov/Traces/study/} and searching the
respective SRA study accession.

@@

\textbf{Figure 2: Treatment Response Based on Models Built for Adenoma,
SRN, or Carcinoma.} A) Positive probability change from initial to
follow up sample in those with adenoma. B) Positive probability change
from initial to follow up sample in those with SRN. C) Positive
probability change from initial to follow up sample in those with
carcinoma..

\textbf{Figure 3: Common OTUs Model Rank Based on Median MDA.} For each
common OTU the lowest taxonomic identification and importance rank for
each model run is shown.

@@

\textbf{Table 1: Demographic Data of Training Cohort}

\textbf{Table 2: Demographic Data of Pre and Post Treatment Cohort}

@@

\hypertarget{ref-sze_looking_2016}{}
28. Sze MA, Schloss PD. Looking for a Signal in the Noise: Revisiting
Obesity and the Microbiome. mBio. 2016;7.

\hypertarget{ref-obrien_impact_2013}{}
29. O'Brien CL, Allison GE, Grimpen F, Pavli P. Impact of colonoscopy
bowel preparation on intestinal microbiota. PloS One. 2013;8:e62815.

>>>>

\section{The effect of treatment on the microbiota of patients diagnosed
with colonic
lesions}\label{the-effect-of-treatment-on-the-microbiota-of-patients-diagnosed-with-colonic-lesions}

\begin{center}
\vspace{25mm}

@@

\textbf{Background.} Colorectal cancer (CRC) is a worldwide health
problem. Despite growing evidence that members of the gut microbiota can
drive tumorigenesis, little is known about what happens to the
microbiota after treatment for an adenoma or carcinoma. This study
tested the hypothesis that treatment for adenoma or carcinoma alters the
abundance of bacterial populations associated with disease to those
associated with a normal colon. We tested this hypothesis by sequencing
the 16S rRNA genes in the feces of 67 individuals before and after
treatment for adenoma (N=22), advanced adenoma (N=19), and carcinoma
(N=26).

\textbf{Results.} There were large changes to the bacterial communities
associated with treatment across the three groups. The communities from
patients with carcinomas changed significantly more than those with
adenoma following treatment (P-value=5.4e-05); however, there was not a
significant difference between those with advanced adenoma and those
with adenoma or carcinoma (P-value \textgreater{} 0.05). Although
treatment brought about large intrapersonal changes, the change in the
abundance of individual OTUs to treatment was not consistent within
diagnosis groups (P-value \textgreater{} 0.05). Because the distribution
of OTUs across patients and diagnosis groups was patchy, we used the
Random Forest machine learning algorithm to identify groups of OTUs that
allowed us to successfully distinguish between pre- and post-treatment
samples for each of the diagnosis groups. However, across the three
models, there was little overlap between the OTUs that were indicative
of treatment. Next, we used a larger cohort that contained individuals
with normal colons and those with adenomas, advanced adenomas, and
carcinomas to determine whether individuals who underwent a treatment
were more likely to have OTUs associated with normal colons. We again
built Random Forest models and measured the change in the positive
probability of having one of the three diagnoses. Although we could
clearly differentiate pre- and post-treatment communities from the three
diagnosis groups, only those patients that initially had carcinomas
experienced a significant decrease in positive probability of having a
carcinoma (P-value \textless{} 0.05). Finally, we tested whether the
type of treatment impacted the microbiota of those diagnosed with
carcinomas and were unable to detect any significant differences in
characteristics of these communities between individuals treated with
surgery alone and those treated with chemotherapy or chemotherapy and
radiation (P-value \textgreater{} 0.05).

\textbf{Conclusions.} Further exploration of the relationship between
diagnosis, treatment, and the impact on the microbiota will yield
improvements in disease management.

@@

Building on this evidence, several human-based studies have identified
unique signatures of colonic lesions {[}14--19{]}. One line of research
has identified community-level differences between those bacteria that
are found on and adjacent to colonic lesions and have supported a role
for \emph{Bacteroides fragilis}, \emph{Escherichia coli}, and
\emph{Fusobacterium nucleatum} in tumorigenesis {[}20--22{]}. Others
have proposed feces-based biomarkers that could be used to diagnose the
presence of colonic adenomas and carcinomas {[}23--25{]}. These studies
have associated \emph{Fusobacterium nucleatum} and other oral pathogens
with lesions (adenoma, advanced adenoma, and carcinoma). They have also
noted that the loss of bacteria generally thought to produce short chain
fatty acids, which can suppress inflammation, is associated with colonic
lesions. Gut bacteria have a role in tumorigenesis and there is promise
that these populations may be useful biomarkers for aiding in the early
detection of disease.

@@

\textbf{\emph{Treatment alters the bacterial community structure of
patients diagnosed with colonic lesions.}} Within our 67-person cohort
we tested whether the microbiota of patients with adenoma (N=22),
advanced adenoma (N=19), or carcinoma (N=26) had any broad differences
between pre- and post-treatment samples {[}Table 1{]}. The structure of
the microbial communities of the pre- and post-treatment samples
differed {[}Figure 1A{]}. We found that the communities obtained pre-
and post-treament among the patients with carcinomas changed
significantly more than those patients with adenoma (P-value = 5.4e-05).
There were no significant differences in the amount of change observed
between the patients with adenoma and advanced adenoma or between the
patients with advanced adenoma and carcinoma (P-value \textgreater{}
0.05). Next, we tested whether there was a consistent direction in the
change in the community structure between the pre and post-treatment
samples for each of the diagnosis groups {[}Figure 1B-D{]}. We only
observed a consistent shift in community structure for the patients with
carcinoma when using a PERMANOVA test (adenoma P-value0.999, advanced
adenoma P-value0.945, and carcinoma P-value0.005). Finally, we measured
the number of observed OTUs, Shannon evenness, and Shannon diversity in
the pre and post-treatment samples and did not observe a significant
change for any of the diagnosis groups (P-value \textgreater{} 0.05)
{[}Table S1{]}.

\textbf{\emph{The effects of treatment are not consistent across
treatment groups.}} We used two approaches to identify those bacterial
populations that change between the two samples for each diagnosis
group. First, we sought to identify individual OTUs that could account
for the change in overall community structure. However, using a paired
Wilcoxon test we were unable to identify any OTUs that were
significantly different in the pre and post-treatment groups {[}Figure
S1{]}. It is likely that high inter-individual variation and the patchy
distribution of OTUs across individuals limited the statistical power of
the test. To overcome these problems we developed Random Forest models
to identify collections of OTUs that would allow us to differentiate
between pre and post-treatment samples from each of the diagnosis
groups. Because of the relatively small number of subjects in each
group, we restricted our models to only incorporate 10 OTUs to limit the
likelihood that that models would overfit the data. Despite this
restriction, the models performed well (adenoma AUC=0.69 - 0.92,
advanced adenoma AUC=0.8 - 1, carcinoma AUC=0.82 - 0.98). Interestingly,
the 10 OTUs that were used for each model had little overlap with each
other {[}Figure 2{]}. These results support the earlier community-wide
analysis where we observed that the treatment had an impact on the
overall community structure; however, the effect of treatment was not
consistent across patients and diagnosis groups.

\textbf{\emph{Post-treatment samples from patients with carcinoma change
towards a microbiota associated with normal}} Next, we determined
whether treatment changed the microbiota in a way that the
post-treatment communities resembled that of patients with normal
colons. To test this, we used an expanded cohort of 423 individuals that
were diagnosed under the same protocol as having normal colons or colons
with adenoma, advanced adenoma, or carcinoma {[}Table 2{]}. We then
constructed Random Forest models to classify the pre and post-treatment
samples as having their original diagnosis or having a normal colon. The
models performed well (adenoma AUC=0.62 - 0.72, advanced adenoma
AUC=0.68 - 0.77, carcinoma AUC=0.84 - 0.9; Figure S2). The OTUs that
were incorporated into the adenoma and advanced adenoma models largely
overlapped and those OTUs that were used to classify the carcinoma
samples were largely distinct from those of the other two models
{[}Figure 3A{]}. Among the OTUs that were shared across the three models
were those populations commonly considered as commensals (e.g.
\emph{Faecalibacterium}, \emph{Lachnospiraceae}, \emph{Bacteroides},
\emph{Dorea}, \emph{Anaerostipes}, and \emph{Roseburia}) {[}Figures
3B{]}. Although many of these OTUs were also included in the model
differentiating between patients with normal colons and those with
carcinoma, this model also included OTUs affiliated with populations
that have previously been associated with carcinoma
(\emph{Fusobacterium}, \emph{Porphyromonas}, \emph{Parvimonas})
{[}Figure S3{]}. Finally, we applied these three models to the pre and
post-treatment samples for each diagnosis group and quantified the
change in the positive probability of the model. A decrease in the
positive probability would indicate that the colon more closely
resembled that of a patient with a normal colon. There was no
significant change in the positive probability for the adenoma or
advanced adenoma groups {[}Figure 4{]}. The positive probability for the
pre and post-treatment samples from patients diagnosed with carcinoma
significantly decreased. In fact, only 6 of the 26 patients (23.08\%)
that were initially diagnosed with a carcinoma had a higher positive
probability after treatment. One of those was re-diagnosed with
carcinoma on the follow up visit. These results indicate that although
there are changes in the microbiota associated with treatment, those
experienced by the patients in the carcinoma group were the only ones
where the change was directed towards what would be found in a normal
colon

@@

This study builds upon previous work that demonstrated a role for the
microbiota in tumorigenesis and that the microbiota contains biomarkers
to diagnosis colonic lesions {[}5--9,23--27{]}. Our study focused on
comparing the microbiota of patients diagnosed with adenoma, advanced
adenoma, and carcinoma before and after treatment. For all three groups
of patients we observed changes in their microbiota. The microbiota of
patients diagnosed with carcinoma changed significantly more than the
other groups. This change resulted in communities that more closely
resembled those of patients with a normal colon. For the patients
diagnosed with carcinomas there was a larger change between their
initial and follow up samples than was observed for patients diagnosed
with either adenoma or advanced adenoma. The changes in the community
structure of those diagnosed with carcinoma were similar and resulted in
communities that more closely resembled those found in patients with
normal colons.

Understanding the effect of treatment on the microbiota of those
initially diagnosed with carcinomas may have important implications into
understanding disease recurrence. The microbiota changes associated with
treatment among patients initially diagnosed with carcinomas resulted in
community structures that resembled patients with normal colons. This
would suggest that treatment for carcinoma is not only successful for
removing the carcinoma but also at removing bacteria associated with
them. Interestingly, 6 of the 26 patients diagnosed with carcinomas had
an elevated probability of having carcinomas after treatment. This leads
us to hypothesize that these individuals may be at a high risk of
recurrence. Interestingly, the 23.08\% prevalence of increased carcinoma
probability from our study is within the expected rate of recurrence
(20-30\% {[}3,4{]}), which would support our hypothesis . Unfortunately,
because of the design of our study, it was not possible to determine
whether individuals who had an increased positive probability for
carcinoma experienced recurrence. Regardless, it is intriguing to think
of using biomarkers from the microbiota to not only predict the presence
of lesions, but to also assess the risk of developing them.

It is interesting that those patients diagnosed with adenoma and
advanced adenoma did not experience a shift towards a community
structure that resembled those with normal colons. This may be due to
the fundamental differences between the features of adenomas and
advanced adenomas and carcinoma. Specifically, carcinomas may create an
inflammatory milieu that would impact the structure of the community and
removal of that stimulus would alter said structure. In addition, it is
possible that the difference between the microbiota of patients with
adenoma and advanced adenoma and those with normal colons is subtle.
This is supported by the reduced ability of our models to correctly
classify patients with adenomas and advanced adenomas relative to those
diagnosed with carcinomas {[}Figure S2{]}. Given the patchy distribution
of microbiota across patients in the different diagnosis groups, it is
possible that we lacked the statistical power to adequately characterize
the change in the communities following treatment. A final hypothesis is
that the specific type of treatment altered the structure of the
microbiome. The treatment to remove adenomas and advanced adenomas was
either polyp removal or surgical resection whereas it was surgical
resection alone or in combination with chemotherapy or with chemotherapy
and radiation for individuals with carcinoma. Because chemotherapy and
radiation target rapidly growing cells, these treatments would be more
likely to cause a turn over of the colonic epithelium driving a more
significant change in the structure of the microbiota. Although, we were
able to test for an effect across these specific types of treatment, the
number of patients in each treatment group was relatively small.

It was not clear why 6 of the 26 patients with carcinoma had a higher
probability of having carcinomas after treatment. We hypothesized that
these individuals may have had more severe tumors; however, the tumor
severity of these 6 individuals (3 with Stage II and 3 with Stage III)
was similar to the distribution observed among the other 20 patients. We
also hypothesized that we may have sampled these patients later than the
rest and their communities may have reverted to a carcinoma-associated
state; however, there was not a statistically significant difference in
the length of time between sample collection among those whose
probabilities increased or decreased (Wilcoxon Test; P-value=0.56).
Finally, it is possible that these patients may not have responded to
treatment as well as the other 20 patients diagnosed with carcinoma and
so the microbiota may not have been impacted the same way. Again,
further studies looking at the role of the microbiota in recurrence are
needed to understand the dynamics following treatment.

This study builds upon existing work that has established a role for the
microbiota in tumorigenesis and that demonstrated the utility of
microbiome-based biomarkers to predict the presence of colonic lesions.
The most exciting future direction from the current study is the
possibility that markers within the microbiota could be used to evaluate
the effect of treatment and predict recurrence for those initially
diagnosed with carcinoma. If such an approach is effective, it might be
possible to even target the microbiota as part of combined therapies
including currently standard approaches. Our data provides additional
evidence on the importance of the microbiota in tumorigenesis by
addressing the recovery of the microbiota after treatment and opens
interesting avenues of research into how these changes may affect
recurrence.

@@

\textbf{\emph{Study Design and Patient Sampling.}} Sampling and design
have been previously reported in Baxter, et al {[}23{]}. Briefly, study
exclusion involved those who had already undergone surgery, radiation,
or chemotherapy, had colorectal cancer before a baseline fecal sample
could be obtained, had IBD, a known hereditary non-polyposis colorectal
cancer, or familial adenomatous polyposis. Samples used to build the
models for prediction were collected either prior to a colonoscopy or
between one and two weeks after initial colonoscopy. The bacterial
community has been shown to normalize back to a pre-colonoscopy
community within this time period {[}28{]}. Our study cohort consisted
of 67 individuals with an initial sample as described and a follow up
sample obtained between 188 - 546 days after treatment of lesion
{[}Table 1{]}. Patients were diagnosed by colonoscopic examination and
histopathological review of biopsies. Patients were classified as having
advanced adenoma if they had an adenoma greater than 1 cm, more than
three adenomas of any size, or an adenoma with villous histology. This
study was approved by the University of Michigan Institutional Review
Board. All study participants provided informed consent and the study
itself conformed to the guidelines set out by the Helsinki Declaration.

\textbf{\emph{16S rRNA Gene Sequencing.}} Sequencing was completed as
described by Kozich, et al. {[}29{]}. DNA extraction used the 96-well
Soil DNA isolation kit (MO BIO Laboratories) and an epMotion 5075
automated pipetting system (Eppendorf). The V4 variable region was
amplified and the resulting product was split between three sequencing
runs with normal, adenoma, and carcinoma evenly represented on each run.
Each group was randomly assigned to avoid biases based on sample
collection location. The pre- and post-treatment samples were sequenced
on the same run.

\textbf{\emph{Sequence Processing.}} The mothur software package
(v1.37.5) was used to process the 16S rRNA gene sequences and has been
previously described {[}29{]}. The general workflow using mothur
included merging paired-end reads into contigs, filtering for low
quality contigs, aligning to the SILVA database {[}reference{]},
screening for chimeras using UCHIME {[}reference{]}, classifying with a
naive Bayesian classifier using the Ribosomal Database Project
(RDP){[}reference-Wang{]}, and clustered into Operational Taxonomic
Units (OTUs) using a 97\% similarity cutoff with an average neighbor
clustering algorithm {[}reference-Westcott\&SchlossAEM{]}. The number of
sequences for each sample was rarefied to 10523 to minimize the impacts
of uneven sampling.

\textbf{\emph{Model Building.}} The Random Forest {[}30{]} algorithm was
used to create the model used to create the three models used. The
adenoma model classified normal versus adenoma, advanced adenoma was
normal versus advanced adenoma, and carcinoma was normal versus
carcinoma. The toal number of individuals in this data set was 423
individuals {[}Table 1{]}. There were a total of 239 individuals in the
adenoma model, 262 individuals in the advanced adenoma model, and 266
indivdiuals in the carcinoma model {[}Table 1{]}. Each model was then
applied to our 67-person cohort testing prediction of adenoma intial
(adenoma n = 22) versus adenoma follow up (adenoma n = 0), advanced
adenoma initial (advanced adenoma n = 19) versus advanced adenoma follow
up (advanced adenoma n = 0), carcinoma initial (carinoma n = 26) versus
carcinoma follow up (carcinoma n = 1).

@@

Assessment of the most important OTUs to the model involved counting the
number of times an OTU was present in the top 10\% of mean decrease in
accuracy (MDA) for each of the 100 different splits run. This was then
followed with filtering of this list to variables that were only present
in more than 50\% of these 100 runs. The final collated list of
variables was then run through the mtry optimization again. Once the
ideal mtry was found the entire sample set specific to normal versus
adenoma, normal versus advanced adenoma, or normal versus carcinoma was
used to create the final Random Forest model on which classifications on
the 67-person cohort was completed. For all three models the final
optimized mtry was 2. The hyper-parameter, mtry, defines the number of
variables to investigate at each split before a new division of the data
was created with the Random Forest model.

\textbf{\emph{Statistical Analysis.}} The R software package (v3.3.2)
was used for all statistical analysis. Comparisons between bacterial
community structure utilized PERMANOVA {[}31{]} in the vegan package
(v2.4.1). Comparisons between probabilities as well as overall OTU
differences between pre- and post-treatment samples utilized a paired
Wilcoxon ranked sum test. Where multiple comparison testing was
appropriate, a Benjamini-Hochberg (BH) correction was applied {[}32{]}
and a corrected P-value of less than 0.05 was considered significant.
Unless otherwise stated the P-values reported are those that were BH
corrected. Model rank importance was determined by obtaining the median
MDA from the 100, 20 repeated 10-fold cross validation and then ranking
from largest to smallest MDA.

\textbf{\emph{Reproducible Methods.}} A detailed and reproducible
description of how the data were processed and analyzed can be found at
\url{https://github.com/SchlossLab/Sze_followUps_2017}. Raw sequences
have been deposited into the NCBI Sequence Read Archive (SRP062005 and
SRP096978) and the necessary metadata can be found at
\url{https://www.ncbi.nlm.nih.gov/Traces/study/} and searching the
respective SRA study accession.

@@

\textbf{Figure 2: The 10 OTUs used to classify treatment for Adenoma,
Advanced Adenoma, and Carcinoma.} A) Adenoma OTUs. B) Advanced Adenoma
OTUs. C) Carcinoma OTUs.

\textbf{Figure 3: Treatment Response Based on Models Built for Adenoma,
Advanced Adenoma, or Carcinoma.} A) Positive probability change from
initial to follow up sample in those with adenoma. B) Positive
probability change from initial to follow up sample in those with
advanced adenoma. C) Positive probability change from initial to follow
up sample in those with carcinoma..

\textbf{Figure 4: Common OTUs to All Models.} A) Venn diagram showing
the OTU overlap between each model. B) For each common OTU the lowest
taxonomic identification and importance rank for each model run is
shown.

@@

\textbf{Table 1: Demographic Data of Pre and Post Treatment Cohort}

\textbf{Table 2: Demographic Data of Training Cohort}

@@

\hypertarget{ref-obrien_impact_2013}{}
28. O'Brien CL, Allison GE, Grimpen F, Pavli P. Impact of colonoscopy
bowel preparation on intestinal microbiota. PloS One. 2013;8:e62815.
